---
title: "`r params$project_title`"
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
params:
  organisation: Karolinska Institutet
  pi_mail: krisina.broliden@ki.se
  pi_name: Kristina Broliden
  project_title: Broliden_5325
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 10,
                      results    = "hold",
                      message    = FALSE,
                      warning    = FALSE)

setwd("/Users/vilkal/work/Brolidens_work/Projects/Gabriella_repo/reports/rmarkdown")
```

#Load libraries and functions

```{r, echo = FALSE}
library(tidyverse)
library(phyloseq)
library(Biostrings)
```

```{r function}
######################## FASTA Function ########################
write.fasta <- function(df, file.out, open = "w"){
  outfile <- file(description = file.out, open = "w")
  write.oneseq <- function(sequence, name){
    writeLines(paste(">", name, sep = ""), outfile)
    writeLines(sequence, outfile)
    return(name)
  }
  map2_chr(df$Sequence, df$Seq_ID, ~write.oneseq(.x, .y))
  close(outfile)
}

```

# Loading data and metadata

```{r Get ASV and tax tables, message=FALSE}
paths_p <- c('../../data/phyloseq_boston_r1.RDS',   # CVL V3
             '../../data/phyloseq_boston_r2.RDS',   # Tissue V3 & CVL V2
             '../../data/phyloseq_Sthlm.RDS')       # CVL V2, only v4 region

n <- c("boston_r1", "boston_r2", "Sthlm")
ps_list <- map(paths_p, ~readRDS(.x)) %>% set_names(n)

seqtabs <- map(ps_list, ~data.frame(.x@otu_table@.Data))
taxtabs <- map(ps_list, ~as_tibble(.x@tax_table@.Data, rownames = 'Sequence'))
sample_info <- map(ps_list, ~as_tibble(sample_data(.x), rownames = 'SampleID')) 
```

## Reshape seqtab 

```{r reshape seqtabs}
seqtab.nochim.reshaped <- map(seqtabs, ~.x %>%
                            rownames_to_column(var = "Sample_ID") %>%
                            pivot_longer(., -Sample_ID) %>%
                            dplyr::rename(Sequence="name", Reads="value")
)
seqtab.taxa <- seqtab.nochim.reshaped %>%
                map2(., taxtabs, ~left_join(.x, .y, by="Sequence"))
```

## Creat Short ID For Sequences across runs

```{r SeqID dict, include=FALSE, results='hide'}
############### Create dictonary for Sequence id ####
Seq_ID_key <-seqtab.taxa %>% 
  bind_rows(.id = "Sequencing_run") %>%
  # Get only uniqe sequences by grouping on all taxonomy
  group_by(Sequence,Kingdom,Phylum,Class,Order,Family,Genus,Species) %>%
  #summarize(n()) %>% filter(`n()`>1) %>% .$`n()` %>% sum(.)
  summarise(Reads_per_taxa = sum(Reads),
            Sequencing_run = paste(unique(Sequencing_run), collapse="; "),
            .groups = "drop") %>%
  bind_cols(Seq_ID = c(paste0("Seq_",sprintf("%05.0f", 1:nrow(.))))) %>%
  {. ->> seqtab.taxa.all } %>% 
  select(Seq_ID, Sequence)

write.csv(Seq_ID_key, file=paste0("../../results/Seq_ID_key.csv"), quote=FALSE, row.names=FALSE)

# Check that all sequences are unique:
#S <- Seq_ID_key %>% group_by(Sequence) %>% summarize(n()) %>% filter(`n()`>1) %>% .$`n()` %>% sum(.)
```

## Create Query files

```{r Querys, echo=T, results='hide'}
# Identify sequences not assigned by RDP at species level
unnasigned_0 <- seqtab.taxa.all %>%
  filter(., is.na(Species)) #%>%
  #filter(., Reads != 0) 

# Create Query files
write.fasta(Seq_ID_key, paste0("../../results/BVAB_Query.fasta"))
write.fasta(unnasigned_0, paste0("../../results/optivag_Query.fasta"))
```

# BLAST
## Download databases

```{r Download data base}
db_dir <- "../../data/database/"
if( isFALSE(dir.exists(db_dir)) ) { dir.create(db_dir,recursive = TRUE) }

url <- c( "https://raw.githubusercontent.com/ctmrbio/BVAB-and-Lac-sequences/master/BVAB_rRNA_database.fa",
          "https://github.com/ctmrbio/optivag/raw/master/database/db/16S/v0.1/optivag_db.fasta.gz",
          "https://github.com/ctmrbio/optivag/raw/master/database/db/16S/v0.1/optivag_seqinfo.csv")

walk(url, ~download.file(.x, paste0(db_dir, basename(.x)), method="auto"))
system("gunzip ../../data/database/*.gz")

OptiVag <- read_csv("../../data/database/optivag_seqinfo.csv", col_names = T)
```

## BLAST seraches

```{r BLASTn search}
# Make uniqe identifyers in the BVAB file
fa = readDNAStringSet(paste0(db_dir,"BVAB_rRNA_database.fa"))
ID <-  make.unique(paste0(str_match(names(fa), "^BVAB._"),"0"), sep = "")
names(fa) = paste(ID, names(fa))
writeXStringSet(fa, "../../data/database/BVAB_db.fasta")

for(name in c("BVAB", "optivag")) { 
  db_dir     = "../../data/database/"
  result_dir = "../../results/"
  if (isFALSE(is_empty(list.files(path = db_dir, pattern = paste0(name,".+\\.nhr"))))) {
    message("using database provided in path:")
    message(paste0(db_dir, name, "_reference.db"))
  } else {
    system2(command = "makeblastdb", 
                          args = c("-in", paste0(db_dir, name, "_db.fasta"), 
                                  "-out", paste0(db_dir, name, "_reference.db"), 
                                  "-parse_seqids", 
                                  "-dbtype", "nucl"),
                         wait = TRUE,
                         stdout = TRUE)
  }
  blast_out <- system2(command = "blastn", 
                       args = c("-db", paste0(db_dir, name, "_reference.db"), 
                                "-query", paste0(result_dir, name, "_Query.fasta"), 
                                #"-out", paste0(result_dir, name, "_hits.tab"),
                                "-outfmt", 6, 
                                "-perc_identity", 99.5,
                                "-qcov_hsp_perc", 90),
                       wait = TRUE,
                       stdout = TRUE) %>%
  as_tibble() %>% 
  separate(col = value, sep = "\t", convert = TRUE,
           into = c("qseqid","sseqid","pident","length","mismatch","gapopen",
                    "qstart","qend","sstart","send","evalue","bitscore") )
  
  write_tsv(blast_out, paste0("../../results/", name, "_hits.tsv"))
}

```

## Load BLAST results

```{r Load BLAST hits, message=FALSE}
blast_OptiVag <- read_table2(paste0("../../results/OptiVag_hits.tsv"))
blast_BVAB <- read_table2(paste0("../../results/BVAB_hits.tsv"))

```

## Retrive Genus and Species hits from BLAST search

```{r Retrive hits}
# Resolve ambiguously asigned genus and Lacto spp.
OptiVag_hits <- blast_OptiVag %>%
                left_join(select(OptiVag, Full_taxonomy, Species, Seq_ID), 
                          by=c(sseqid='Seq_ID')) %>%
                select(Seq_ID=qseqid, Full_taxonomy, Species, Ref_iD=sseqid, 
                       Identity=pident, Mismatch=mismatch) %>%
                group_by(Seq_ID) %>% 
# removes hits with mismatch=1, for sequences with multiple hits where on of the hits has mismatch=0 
                filter(!(n()>1 & any(grepl('0', Mismatch)) & Mismatch == 1)) %>%
                ungroup() %>%
                separate(Species, into=c('Genus', 'Species'), sep = '_') %>% 
                separate(Full_taxonomy, into=c("Full_taxonomy", NA), sep = '(;)(?:.(?!;))+$') %>%
                select(Seq_ID, Full_taxonomy, Genus, Species) %>%
                unique() %>%
# the last group is dropped after summarize (.groups argument)
                group_by(Seq_ID, Full_taxonomy, Genus) %>% 
                summarise(Species = paste(Species, collapse="/"), 
                          .groups='drop_last') %>%
                mutate(Genus = paste(Genus, collapse="/")) %>%
                left_join(select(unnasigned_0, Seq_ID, Reads_per_taxa), 
                          by='Seq_ID') %>%
                ungroup() 

BVAB_hits <- blast_BVAB %>%
              select(qseqid, sseqid) %>%
              left_join(select(seqtab.taxa.all, Seq_ID, Reads_per_taxa),
                          by=c('qseqid'='Seq_ID')) %>%
              separate(sseqid, c('Genus', 'Species')) %>%
              dplyr::rename(Seq_ID=qseqid) %>%
              select(Seq_ID, Genus, Reads_per_taxa) %>%
              unique() 

all_hits <- bind_rows(BVAB_hits, OptiVag_hits) %>%
              unique()
```

## BLAST result info

```{r Information about BLAST result}
print(paste0("Number of sequences annotated by OptiVag:  ", length(unique(OptiVag_hits$Seq_ID))))

print(paste0("Number of sequences annotated by BVAB:     ", length(unique(BVAB_hits$Seq_ID))))
 
overlap <- intersect(BVAB_hits$Seq_ID, OptiVag_hits$Seq_ID)
ifelse(length(unique(all_hits$Seq_ID)) == length(all_hits$Seq_ID), "There is no duplicate species annotation", paste0(message("Resolving overlapping annotation between BVAB and OptiVag:"), paste(overlap,  collapse=', ')))

doNextChunk <- isFALSE(length(unique(all_hits$Seq_ID)) == length(all_hits$Seq_ID))

```

# Resolve Taxonomy
## Resolve overlapping annotation between BVAB and OptiVag

```{r Resolve Overlap, eval = doNextChunk}
# Identify which annotation that overlap for BVAB sequences 
duplicate <- all_hits %>%
  group_by(Seq_ID) %>% 
  filter(n()>1) %>%
  filter((grepl('BVAB', Genus))) %>%
  ungroup()

# Filter out duplicated annotation for BVAB
filtered_hits <- all_hits %>%
  filter(!(Seq_ID %in% duplicate$Seq_ID & !(grepl('BVAB', Genus)))) 

# is there duplicates other than BVAB?
d <- filtered_hits %>% group_by(Seq_ID) %>% summarize(n(), .groups='drop_last') %>% filter(`n()`>1)  %>% .$Seq_ID

# prints message about duplicate status
ifelse(length(unique(filtered_hits$Seq_ID)) == length(filtered_hits$Seq_ID), "Overlaping annotation resolved", paste0(message("Resolving duplicate taxonomy asignment:"), paste(d,  collapse=', ')))

# If duplicates are identified, run next chunk
doNextChunk <- isTRUE(length(d)>=1) 

```

## Resolve duplicate taxonomy

```{r Resolve duplicate taxonomy, eval = doNextChunk}
# Collapse species annotation for sequences with several hits
filtered_2 <- aggregate(. ~Seq_ID, data = filtered_hits, na.action=NULL, FUN=function(x) { paste(unique(x), collapse = '/') }) 

# Sequence ids that are duplicate:
dd <- filtered_2 %>% group_by(Seq_ID) %>% summarize(n(), .groups='drop_last') %>% filter(`n()`>1)  %>% .$Seq_ID

# Message to terminal
ifelse(length(unique(filtered_hits$Seq_ID)) == length(filtered_2$Seq_ID), "Duplicate annotation resolved", paste0(message("WARNING! still duplicate annotation:"), paste(dd,  collapse=', ')))

filtered_hits <- filtered_2
```

## Merge New Annotation With Tax Table

```{r Save updated tax table}
#### Update taxa ####
update_taxa <- seqtab.taxa.all %>%
  left_join(., select(filtered_hits, Seq_ID, Genus_u="Genus",
                      Species_u="Species", Full_taxonomy), by='Seq_ID') %>%
  mutate(identical = ifelse(str_detect(.$Genus_u, "BVAB")|
                                     .$Genus_u == .$Genus|
                                           is.na(.$Genus), "yes", "no"))

new_taxtable <- update_taxa %>%
  mutate(Genus_f = ifelse(.$identical=="no"|is.na(.$identical), 
                          .$Genus, .$Genus_u)) %>%
  mutate(Species_f = ifelse(.$identical=="no"|is.na(.$identical), 
                            .$Species, .$Species_u)) %>%
  select(-Genus, -Species, -Genus_u, -Species_u, -identical, -Sequencing_run) %>%
  dplyr::rename(Genus="Genus_f", Species="Species_f") %>%
  select(Seq_ID, everything(), -Full_taxonomy, -Reads_per_taxa)

write.csv(new_taxtable, file=paste0("../../data/SeqID_to_taxa.csv"), quote=FALSE, row.names=FALSE)
```

## Report of unasigned sequences

```{r Saving unnasigned sequences}
#### Taxa that still have no Genus or Lacto spp. annotation ####
unnasigned_final <- new_taxtable %>% 
  left_join(select(seqtab.taxa.all, Sequence, Reads_per_taxa, Sequencing_run), by="Sequence") %>%
  filter(is.na(.$Genus)) %>%
  arrange(desc(Reads_per_taxa)) %>%
  arrange(desc(Family)) %>%
  filter(Reads_per_taxa >100)

write.csv(unnasigned_final, 
          file=paste0("../../results/Unnasigned_tax_report.csv"), quote=FALSE, row.names=FALSE)

unnasigned_final
```

# Save results

## merge seqtab with new taxonomy

```{r Create ASV tables}
# Add sample info and updated taxonomy 
ASV_wCtrl <- seqtab.nochim.reshaped %>% 
            map2(., sample_info, 
                 ~left_join(.x, select(.y, SampleID, ID, Sample_type, Visit, Place_sequenced,
                                            Description, Include, Include_final),
                            by=c("Sample_ID"="SampleID")) ) %>%
                  {. ->> ASV } %>%
            map(., ~ .x %>%
                  select(Sequence, Sample_ID, Reads) %>%
                  pivot_wider(., names_from=Sample_ID, values_from=Reads) %>%
                  left_join(., new_taxtable, by="Sequence")
                  )

#### ASV with all controls and replicate samples removed ####
ASV_names <- c("ASV_CVL_V2_B1", "ASV_CVL_V3", "ASV_tissue_B1", 
               "ASV_CVL_V2_B2", "ASV_tissue_B2", "ASV_CVL_V2_C" )
ASV_list <- map(ASV, ~ .x %>%
                  filter(grepl("yes", .$Include)) %>%
                  group_split(Sample_type, Visit, .keep = T)
            ) %>%
            flatten() %>%
            set_names(ASV_names) %>%
            map(., ~ .x %>%
                select(Sequence, ID, Reads) %>%
                pivot_wider(., ,names_from=ID, values_from=Reads) %>%
                left_join(., new_taxtable, by="Sequence") 
                )

```

## Save csv

```{r Save ASV files}
imap(ASV_list, 
     ~write.csv(.x, file=paste0("../../data/",.y,".csv"), quote=FALSE, row.names=F) )

```