---
title: "dada2 Single-End Protocol - `r params$run`"
author: "Seth M. Bloom"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
  word_document: default
params:
  run: "Boston_run1"
---

### Researchers/collaborators:  
* Gabriella Edfeldt <gabriella.edfeldt@ki.se>  
* Jiawu Xu <jxu25@mgjh.harvard.edu>  
* Douglas S. Kwon <dkwon@mgh.harvard.edu>  
  
  
## References:  
* Callahan, B. J., McMurdie, P. J., Rosen, M. J., Han, A. W., Johnson, A. J. A., & Holmes, S. P. (2016). DADA2: High-resolution sample inference from Illumina amplicon data. Nature Methods, 13(7), 581–3.  
    + https://doi.org/10.1038/nmeth.3869  
* McMurdie, P. J., & Holmes, S. (2013). phyloseq: an R package for reproducible interactive analysis and graphics of microbiome census data. PloS One, 8(4), e61217.  
    + https://doi.org/10.1371/journal.pone.0061217  
* DADA2 tutorial:           http://benjjneb.github.io/dada2/tutorial.html  
* DADA2 big data tutorial:  http://benjjneb.github.io/dada2/bigdata.html  
* Phyloseq:                 https://joey711.github.io/phyloseq/index.html  
* Qiime 1 Mapping files:  http://qiime.org/documentation/file_formats.html#mapping-file-overview  

  

### Description of project and source of data  
  
DESCRIPTION OF PROJECT(S) INCLUDED IN SEQUENCING RUN: 
  
* Project 1:  
    + Researchers:  Gabriella Edfeldt, Jiawu Xu   
    + Number of Samples:  201 + ctrl.
* TECHNICAL CONTROLS    
    + Zymo DNA Standard: 0   
    + Zymo Community DNA Standard: 0    
    + Water_pcrneg: 2    
  

* Sequence files were demultiplexed and split on sample name using Qiime1.  
* This script assigns taxonomy using a variety of different 16S gene reference databases including RDP, GreenGenes, Silva, and HitDB.  
* For archiving purposes, the demultiplexed and split files are stored on O2 in /n/groups/kwon/data1/sequencing_run_archive_2019_02_25_MiSeq/Demultiplexed 
  
    
### Sequencing run parameters and metrics (Boston run 1 only) 
**RUN PARAMETERS (from RunParameters.xml MiSeq output file)**  
   
* Sequencing Date:          **2019_02_25**  
* Run ID:                   **190225M05907 0033 000000000-C8CH9 **   
* Sequencer:                **Instrument M05907**   
* Run type:                 **Single-end V2 300x1 kit (e.g. Single-end V2 300x1 kit)**   
* Flow cell information:    **Serial number: 000000000-C8CH9 Exp.Date: 2019_10_26 **   
* PR2 Buffer information:   **Serial number: MS7625117-00PR2 Exp.Date: 2019_11_04**   
* Reagent kit information:  **Serial number: MS7731647-300V2 Exp.Date: 2019_10_05**  
  
**RUN METRICS (From Illumina BaseSpace record of the run, https://basespace.illumina.com/home/index )**  
  
* Flow cell status:           **QC passed**   
* Target PhiX conc.:          **10%**   
* Read 1 aligned (%):         **4.87%**   
* Read 2 (I) aligned (%)      **%**   
* Clusters passing filter:    **75.16% +/- %**  
* Read 1 % >=Q30:             **82.52%**   
* Read 2 (I) % >=Q30:         **76.26%**   
* Read 1 yield:               **4.85Gbp**   
* Read 2 (I) yield:           **178.59Mbp**   
* Read 1 error rate:          **1.84% +/-0.08%**   
* Read 2 error rate:          **0.00%**   
* Reads passing filter:       **16,235,386**   
* Cluster density:            **1072 +/- 145k/mm^2 (MiSeq target is 865 to 965)**   
* Tiles:                      **28**   
* Legacy Phas/Prephas Read 1: **0.078  /0.154**   
* Read 1 intensity:           **18 +/-0**   
* Read 2 (I) intensity:       **524 +/- 86**      


## Expected output  
1.  A knitted R markdown document (default format is HTML, but can also select options including PDF or MSWord) that contains all comments, code, and figures produced. Key highlights include:  
    + The document returns the names of any samples that failed to pass filter and were eliminated from the final output tables and phyloseq output.  
    + The document calculates and returns cumulative processing time at various steps along the analysis.  
2.  Filtered reads with quality information for each sample stored in newly created subdirectory /Data/filtered.  
    + If sequence files were originally named **samplename**.fastq, the corresponding filtered files will be named **samplename**.filt.fastq  
3.  A newly created subdirectory named /Output containing:  
    + Plot of aggregated read quality scores from the {plot-unfiltered-run-quality} chunk  
    + Plot(s) of read quality scores for individual sample(s) (if code for these plots is included in the {plot-unfiltered-run-quality} chunk)  
    + Plot of error rates from the {learn-errors} chunk  
    + Table of read counts and percentages retained after each sequence analysis step for each sample, saved in a .txt file in TSV format (can be opened in Excel)  
    + Plot of read percentages retained after each sequence analysis step for each sample  
4.  A newly created subdirectory named /ps_objects containing a different phyloseq object for each taxonomic database used. Each phyloseq object is created in the {create-phyloseq} chunk and contains:  
    + An otu_table (ASV table) based on amplicon sequence variants (ASVs) after dada sequence inference and chimera removal  
    + A tax_table based on taxonomic assignments performed in the {assign-taxonomy} chunk
    + sample_data derived from the mapping file.  
    + Phylogenetic trees are not constructed because depending on the type of sample sequenced, they may not be relevant (e.g. in vitro synthetic mixtures of laboratory bacterial strains for QC)  
5. A newly created subdirectory named /RDS containing:  
    + An RDS file containing the sequence table after chimera removal.  
    + An RData workspaced imaging saving the final workspace for the analysis.  
  

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = getwd())
```

```{r start-tracking-processing-time, results = "hold"}
#Define variable with start time of running script.
start_time <- proc.time()
```

```{r initiate-environment, message=FALSE, collapse=TRUE}
# Load libraries
library(ShortRead)
packageVersion("ShortRead")
library(dada2)
packageVersion("dada2")
library("phangorn")
packageVersion("phangorn")
library("phyloseq")
packageVersion("phyloseq")
library("ips")
packageVersion("ips")
library("tidyverse")
packageVersion("tidyverse")

# Set random seed for reproducibility purposes
set.seed(100)

```

```{r define-sequencing-run-date-and-version, results = "hold"}
if(params$run=="Boston_run1"){
  run.name <- "Boston_run1"
  run.date <-  "2019_02_25_v1" 
  mapping_file <- "Mapping_info_Boston_run1.csv"
  # Trim read lenght left and right
  FwdTrimLeft <- 10   
  FwdTrimRight <- 225  
}

if(params$run=="Boston_run2"){
  run.name <- "Boston_run2"
  run.date <-  "2019_05_25" 
  mapping_file <- "Mapping_info_Boston_run2.csv"
  # Trim read lenght left and right
  FwdTrimLeft <- 0   
  FwdTrimRight <- 252  
}

# Sequencing run date in format YYYY_MM_DD
print("Sequencing run date and version:")
run.date
```

```{r Donload data from ENA}
# if(params$run=="Boston_run1"){
#   url <- c(
#     "ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR919/001/SRR9198521/SRR9198521.fastq.gz",
#     "ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR919/007/SRR9198517/SRR9198517.fastq.gz",
#     "ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR919/008/SRR9198518/SRR9198518.fastq.gz"
#     ) }
# dir <- "../data/Fastq"
# dir.create(dir)
# purrr::walk(url, ~download.file(.x, file.path("../data/Fastq/", basename(.x)), method="auto"))
# system("gunzip ../data/Fastq/*.gz")
# 

if(params$run == "Boston_run1"){dir  <- "/Users/vilkal/Raw_data/Microbiome/Run1_Boston_CVLv3_March2019"}
if(params$run == "Boston_run2"){dir <- "/Users/vilkal/Raw_data/Microbiome/Run2_Boston_Tissuev3_June2019"}

files <- list.files(path = dir, pattern =".fastq$", full.names = T, recursive = T)


```

```{r eval=FALSE, include=FALSE}
ENA_info <- read_csv("/Users/vilkal/work/Brolidens_work/Projects/Microbiom_PSC/data/samples-2022-01-30T12 42 43.csv")

mapping_br1 <- readxl::read_xlsx("/Users/vilkal/work/Brolidens_work/Projects/Run1_Boston_CVLv3_March2019/Mapping/Mapping_info_Boston_run1.xlsx")

ENA <- ENA_info %>%
  select(id, SampleID="alias", secondaryId) %>%
  left_join(., mapping_br1, by="SampleID") %>%
  na.omit()
intersect(mapping_br1$SampleID, ENA$SampleID)


# Update again:
# write_csv(ENA, "/Users/vilkal/work/Brolidens_work/Projects/Gabriella_repo/data/Mapping/Mapping_info_Boston_run1.csv")


# ENA_run1 <- read_tsv("/Users/vilkal/work/Brolidens_work/Projects/Microbiom_PSC/results/ENA_files_submission/2022-01-30/ENA_Run_template_Run1.tsv")
# ENA_run2 <- read_tsv("/Users/vilkal/work/Brolidens_work/Projects/Microbiom_PSC/results/ENA_files_submission/2022-01-30/ENA_Run_template_Run2.tsv")
# ENA_T <-read_tsv("/Users/vilkal/work/Brolidens_work/Projects/Microbiom_PSC/results/ENA_files_submission/2022-01-30/ENA_Sample_template_T.tsv")
# ENA_L <-read_tsv("/Users/vilkal/work/Brolidens_work/Projects/Microbiom_PSC/results/ENA_files_submission/2022-01-30/ENA_Sample_template_T.tsv")
# 
# ENA_samples_report <- read_csv("/Users/vilkal/work/Brolidens_work/Projects/Microbiom_PSC/data/samples-2022-01-30T12 42 43.csv")
# 
# ENA_LT <- bind_rows(ENA_L, ENA_T)
# ENA <- bind_rows(ENA_run1, ENA_run2) %>%
#   rename(id="sample") %>%
#   left_join(., select(ENA_samples_report, id, secondaryId, alias), by="id") %>%
#   select(id, study, forward_file_name)
# 
# mapping_1 <- read_csv(file ="../data/Mapping/Mapping_info_Boston_run1.csv", col_names = TRUE) %>%
#   left_join(., ENA, by="id") %>%
#   select(id, study, everything(),-PatID_correct)
# 
# mapping_2 <- read_csv(file ="../data/Mapping/Mapping_info_Boston_run2.csv", col_names = TRUE) %>%
#   left_join(., ENA, by="id") %>%
#   select(id, study, everything(),-PatID_correct)

# cahnge again
# mapping_1 <- read_csv(file ="../data/Mapping/Mapping_info_Boston_run1.csv", col_names = TRUE) %>%
#   mutate(SampleID = str_replace(.$forward_file_name, ".fastq","")) %>%
#   select(-forward_file_name)
# 
# mapping_2 <- read_csv(file ="../data/Mapping/Mapping_info_Boston_run2.csv", col_names = TRUE) %>%
#   mutate(SampleID = str_replace(.$forward_file_name, ".fastq","")) %>%
#   select(-forward_file_name)
# 
#  write_csv(mapping_1, "../data/Mapping/Mapping_info_Boston_run1.csv")
#  write_csv(mapping_2, "../data/Mapping/Mapping_info_Boston_run2.csv")
```

### Filter and trim reads

```{r establish-directory-structure, results = "hold", warning=FALSE}
#Define and print path to working directory
pathwd <- getwd()
print("Working directory:")
pathwd

# Define path to subdirectory containing demultiplexed forward-read fastq files
pathF <- file.path("../data/Fastq")
print("Directory containing raw demultiplexed .fastq sequence files:")
pathF

#Define path to directory containing 16S databases
print("Path to 16S databases:")
pathdb <- file.path("../Resources/RDP_database")
pathdb

#Define path to directory containing mapping file
print("Path to mapping file:")
pathMap <- file.path("../data/Mapping")
pathMap

#Create and define path to subdirectory in which filtered files will be stored
filtpathF <- file.path("../data/Fastq/filtered") 
print("Directory that will contain filtered .fastq files:")
filtpathF
dir.create(filtpathF)

#Create and define path to subdirectory in which to store output figures and tables
print("Directory in which to store output figures and tables:")
pathOut <- file.path("../results/00_dada2_output")
pathOut
dir.create(pathOut)

#Create and define path to subdirectory in which RDS files will be stored
print("Path to directory for saving RDS files:")
pathF.RDS <- file.path(pathOut, "RDS")
pathF.RDS
dir.create(pathF.RDS)

#Create and define path to subdirectory in which RDS files containing phyloseq objects will be stored
print("Path to sub-directory in which to save phyloseq objects:")
pathps0 <- file.path(pathOut, "ps_objects")
pathps0
dir.create(pathps0)
```


```{r Define-sequence-file-names-and-sample-names, results = "hold"}
# File parsing
#Create vector of fastq files in pathF.directory
#If analyzing paired end reads, the search pattern below would need to be modified (see DADA2 tutorial)

print("Sequence files to be analyzed:")
fnFs <- sort(files)
head(fnFs)

#Extract sample names, assuming filenames have format: SAMPLENAME.fastq or SAMPLENAME.fastq.gz
sample.names <- sapply(strsplit(basename(fnFs), ".fast"), `[`, 1)
head(sample.names)
```



```{r plot-unfiltered-run-quality, results = "hold"}
#Plot forward read quality aggregated across all forward read files
p.qual.f.1.agg <- plotQualityProfile(fnFs, aggregate = TRUE) + 
  ggtitle("Fwd read quality aggregated across samples")
p.qual.f.1.agg

#Save figure to file
ggsave(filename = file.path(pathOut, paste0("Read_quality_aggregate_fwd_", run.date, ".pdf")), 
       plot = p.qual.f.1.agg, device = "pdf", width = 8, height = 6, units = "in") 

```


```{r calculate-quality-plot-processing-time, results = "hold"}
#Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```

## Define positions at which to trim filtered data


```{r define-trim-positions, results = "hold"}
## ## ## ## ## ## ## ## ## ##
## USER INPUT REQUIRED HERE ##
## ## ## ## ## ## ## ## ## ##

# Define base positions at which to trim the filtered forward read sequences
# Decisions about where to trim should be based on the quality profile above.
# See DADA2 tutorial for additional information.
# The two variable defined here will then be supplied to the filterAndTrim function
  # FwdTrimLeft gives the posion from the left margin at which to trim
  # FwdTrimRight gives the position from the right margin at which to trim
# For example if FwdTrimLeft is 10 and FwdTrimRight is 240, sequence lengths will be 230 bp.

print(paste0("Trimming forward reads on left at base position ", FwdTrimLeft, 
       " and on right at base position ", FwdTrimRight))
```


## Quality filter and trim
```{r quality-filter-and-trim, results = "hold"}
# Create vector of modified names of fastq files after filtering in
# which the modified file name changes from "samplename.fastq" 
# to "samplename.filt.fastq"

filtFs <- file.path(filtpathF, str_replace(basename(fnFs), "(.fastq)", ".filt\\1"))
head(filtFs)

# Filter and trim sequences listed in fnFs and store them in filtFs, 
# while storing summary in dataframe out

out <- filterAndTrim(fwd = fnFs, filt = filtFs, rev = NULL, filt.rev = NULL,
              trimLeft = FwdTrimLeft, truncLen = c(FwdTrimRight), maxEE = 2, truncQ = 11, maxN = 0, 
              rm.phix = TRUE, compress = TRUE, verbose = TRUE, multithread = TRUE)
head(out)
#truncate file extension (".fastq" or ".fastq.gz") from rownames
rownames(out) <- rownames(out) %>% str_replace(".fast.+", "")
head(out)
```

```{r calculate-filter-trim-processing-time, results = "hold"}
# Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```


## Sequence inference

```{r sample-inference-1, results = "hold"}
## Set paramaters
# Create character of all files in directory filtpathF.1 (i.e. all filtered files)
# with names including string "fastq" (i.e. all *.fastq and *.fastq.gz files)
filtFs.pass <- list.files(filtpathF, pattern="fastq", full.names = TRUE)
head(filtFs.pass)

# Create character vector of sample names from the file names in filtFs.1
# This code assumes all filename = "samplename.filt.fastq"" or "samplename.filt.fastq.gz""
sample.names.filt.pass <- sapply(strsplit(basename(filtFs.pass), ".filt.fast"), `[`, 1) 
head(sample.names.filt.pass)

# Name the elements in filtFs according to the elements in sample.names.filt
names(filtFs.pass) <- sample.names.filt.pass

# Samples for which no sequences passed the filtering step
print("Samples for which no sequences passed filter criteria (i.e. samples excluded from filtered dataset):")
sample.names.filt.fail <- setdiff(sample.names, sample.names.filt.pass)
sample.names.filt.fail

```


```{r learn-errors, results = "hold"}
# Set random seed for reproducibility purposes
set.seed(100)

# Learn forward read error rates
errF <- learnErrors(filtFs.pass, nread=1e6, multithread=TRUE, randomize = "TRUE")

# Plot convergence of error rate computation
plot(dada2:::checkConvergence(errF), type = "o", col = "firebrick3", main = "Convergence")

# Plot calculated errors
p.err.F <- plotErrors(errF, nominalQ = TRUE)
p.err.F

# Save figure to file
ggsave(filename = file.path(pathOut, paste0("Error_rates_fwd_", run.date, ".pdf")),
       plot = p.err.F, device = "pdf", width = 8, height = 6, units = "in")
```

```{r calculate-learn-error-processing-time, results = "hold"}
#Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```

```{r dereplication-and-sample-inference, results = "hold"}
# Dereplicate and apply error rate to resolve sequence variants

## ## ## ## ## ## ## ## ## ## ## ##
## This chunk gives two different options for analyzing the data, depending on the size of the dataset
## and availability of memory for analysis. Use only Option 1 or Option 2 for analysis, not both.
## Comment (or uncomment) the relevant option by highlighting and hitting <command> + <shift> + C
## ## ## ## ## ## ## ## ## ## ## ##

## ## ## ## ## ## ## ## ## ## ## ## ## ## ##
## OPTION 1: **************************** ##
## ## ## ## ## ## ## ## ## ## ## ## ## ## ##

### Code for smaller datasets (e.g. most MiSeq) for which memory is adequate for parallelization ###
### This code analyzes the samples in parallel, which is more efficient but more memory-intensive ###
### Derived from https://benjjneb.github.io/dada2/tutorial.html ###

# Dereplicate the filtered forward sequences by sample
derepFs <- derepFastq(filtFs.pass, verbose = FALSE)

# Set random seed for reproducibility purposes
set.seed(100)

# Run dada on the dereplicated forward sequences to resolve sequence variants
# based on the forward read error model calculated in errF
dadaFs <- dada(derepFs, err = errF, multithread = TRUE)


# ## ## ## ## ## ## ## ## ## ## ## ## ## ## ##
# ## OPTION 2: *************************** ##
# ## ## ## ## ## ## ## ## ## ## ## ## ## ## ##

# ### Code for very large (e.g. HiSeq) datasets if parallelization would exceed available memory ###
# ### This code analyzes the samples in series, which is less efficient but less memory-intensive ###
# ### Derived from https://benjjneb.github.io/dada2/bigdata.html ###
# 
# #Create lists derepFs and dadaFs of same length and element names as sample.names.filt.pass
# derepFs <- vector("list", length(sample.names.filt.pass))
# names(derepFs) <- sample.names.filt.pass
# dadaFs <- vector("list", length(sample.names.filt.pass))
# names(dadaFs) <- sample.names.filt.pass
# 
# #For each sequence file in filtFs.1, dereplicate the sequences from the file
# #then run dada to resolve sequence variants based on the error model errF.1
# #store results in the corresponding named element of singles.1
# for(sam in sample.names.filt.pass) {
#   cat("Processing:", sam, "\n")
#   derepFs[[sam]] <- derepFastq(filtFs.pass[[sam]])
#   # Set random seed for reproducibility purposes
#   set.seed(100)
#   dadaFs[[sam]] <- dada(derepFs[[sam]], err=errF, multithread=TRUE)
# }
# rm(sam)
```
  
```{r calculate-sequence-inference-processing-time}
# Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```

## Construct sequence table and remove chimeras

```{r seqtab-chimera-removal-1, results = "hold"}
# Make sequence table and review dimensions
seqtab.prechimeraremoval <- makeSequenceTable(dadaFs)
print("Sequence table dimensions (samples, resolved sequences) before chimera removal:")
dim(seqtab.prechimeraremoval)

# Inspect distribution of sequence lengths
print("Distribution of sequence lengths before chimera removal:")
table(nchar(getSequences(seqtab.prechimeraremoval)))

# Remove chimeras
seqtab.nochim <- removeBimeraDenovo(seqtab.prechimeraremoval,
                                    method = "consensus",
                                    multithread = TRUE,
                                    verbose = TRUE)

print("Sequence table dimensions (samples, resolved sequences) after chimera removal:")
dim(seqtab.nochim)

# Inspect distribution of sequence lengths
print("Distribution of sequence lengths after chimera removal:")
table(nchar(getSequences(seqtab.nochim)))
print("Proportion of reads eliminated by chimera removal:")
sum(seqtab.nochim) / sum(seqtab.prechimeraremoval)
```

```{r save-RDS-and-image, results = "hold"}
# Save seqtab.nochim RDS file
print("Save seqtab.nochim as RDS file named:")
f.seqtab.nochim.RDS <- file.path(pathF.RDS, paste0("MiSeq_", run.date, "_preprocessing_single_nochim.RDS"))
f.seqtab.nochim.RDS
saveRDS(seqtab.nochim, f.seqtab.nochim.RDS)

```

```{r calculate-chimera-removal-processing-time, results = "hold"}
# Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```

## Track reads through pipeline
The code in the chunk below is substantially altered from the code suggested in the DADA2 tutorial because  
the tutorial code can lead to errors if any samples are eliminated at a step (e.g. if no reads pass filter).  

```{r track-reads}
# Function to calculate reads for each element in dada output
getN <- function(x) sum(getUniques(x))

# Read count remaining at each step of processing. Would need additional columns if analyzing
# paired-end data.

# Create interim dataframes listing the sequence count after each step of processing per sample
print("Read count remaining at each step of processing:")
outdf <- out %>% 
  as.data.frame() %>%
  tibble::rownames_to_column() %>% 
  dplyr::rename(Sample = "rowname", input = "reads.in", filtered = "reads.out")

dadacountdf <- sapply(dadaFs, getN) %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column() %>%
  dplyr::rename(Sample = "rowname", denoisedF = ".")

nonchimdf <- rowSums(seqtab.nochim) %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column() %>%
  dplyr::rename(Sample = "rowname", nonchim = ".")

# Join dataframes by Sample column to track counts for each sample at each step of processing
track <- dplyr::left_join(outdf, dadacountdf, by = c("Sample" = "Sample")) %>%
  dplyr::left_join(nonchimdf, by = c("Sample" = "Sample"))
head(track)

# Percentage of original unfiltered reads remaining at each step of processing
print("Percentage of original unfiltered foward reads remaining at each step of processing:")
track.percent <- dplyr::mutate_at(track, vars(-matches("Sample")), funs(. * 100 / track[["input"]]))
track.percent <- rename_at(track.percent, vars(-matches("Sample")), funs(paste0(., "_percent")))
head(track.percent)

# Create a "long" form version of track.percent for use with plotting
track.percent.long <- track.percent %>% 
  tidyr::gather(key = "Step", value = "Percent", -Sample) 

# Modify track.percent.long so that Step column is a factor with levels in the correct order
track.percent.long[["Step"]] <- track.percent.long[["Step"]] %>% forcats::fct_relevel(colnames(track.percent[-1]))

# Plot percentages of input forward reads remaining after each step on a per sample basis.
p.track <- ggplot(track.percent.long, aes(x = Step, y = Percent)) + 
  geom_line(aes(group = Sample, colour = Sample)) +
  geom_point() +
  labs(title = "Percent of input forward reads remaining\nafter each processing step per sample",
       subtitle = paste0("Fwd read left trim position: ", FwdTrimLeft, "\nFwd read right trim position: ", FwdTrimRight),
       x = "Processing Step", y = "% of Input Reads Remaining") +
  theme(plot.title = element_text(hjust = 0.5, size = 18), 
        plot.subtitle = element_text(hjust = 0.5, size = 16),
        legend.position = "none",
        axis.text = element_text(size = 12), axis.title = element_text(size = 16))
p.track

# Save figure to file
ggsave(filename = file.path(pathOut, paste0("read_track_fwd_L", FwdTrimLeft, "_R", FwdTrimRight, "_", run.date, ".pdf")), 
       plot = p.track, device = "pdf", width = 8, height = 6, units = "in")

# Merge track and track.percent dataframes
track.merged <- dplyr::left_join(track, track.percent, by = c("Sample" = "Sample"))
head(track.merged)

# Save log file of tracking reads to file.
print("Save track.merged dataframe tracking read counts as TSV file named:")
f.track.merged <- file.path(pathOut, paste0("MiSeq_", run.date, "_fwd_read_tracking_log.txt"))
head(f.track.merged)
readr::write_tsv(track.merged, f.track.merged)

```


## Assign taxonomy

```{r assign-taxonomy, results = "hold"}
# RDP
# Record start time for this taxonomy assignment step
tax_start_time = proc.time()
print(paste0("RDP training database: ", "rdp_train_set_16.fa.gz"))
taxa.rdp <- assignTaxonomy(seqtab.nochim, file.path(pathdb, "rdp_train_set_16.fa.gz"), multithread = TRUE) 

# Add RDP species assignment
print(paste0("RDP species assignment database: ", "rdp_species_assignment_16.fa.gz"))
taxa.rdp.plus <- addSpecies(taxa.rdp, file.path(pathdb, "rdp_species_assignment_16.fa.gz"))
writeLines("RDP taxonomy assignment step processing time (seconds):")
proc.time() - tax_start_time

```


```{r calculate-taxonomy-assignment-processing-time, results = "hold"}
# Calculate cumulative processing time
print("Cumulative processing time (seconds):")
proc.time() - start_time
```

## Generate PhyloSeq Objects and final save

```{r create-phyloseq, echo=FALSE, results = "hold"}
# Create PhyloSeq objects

# Load mapping file for sequencing run
print("Sequencing run mapping file:")
f.mapping <- file.path(pathMap, list.files(path = pathMap, pattern = mapping_file))
f.mapping
mapping <- readr::read_csv(file = f.mapping, col_names = TRUE)

names <- mapping$`SampleID`
# Remove any rows with "#SampleID" column value NA and convert to dataframe
mapping <- mapping   %>%
  dplyr::filter(!is.na(`SampleID`)) %>%
  as.data.frame(mapping)

# Convert mapping$#SampleID column to rownames
mapping <- tibble::column_to_rownames(mapping, var = "SampleID")


# RDP
ps0.rdp <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = FALSE), 
                    tax_table(taxa.rdp.plus),
                    sample_data(mapping))
ps0.rdp

# Rename samples
sample_names(ps0.rdp) <- sample_data(ps0.rdp)[["Sample_ID"]]

# Sanity checks
get_taxa_unique(ps0.rdp, "Phylum")

# Save RDS files for downstream analysis
saveRDS(ps0.rdp, file = file.path(pathps0, paste0("phyloseq_",run.name, ".RDS")))

# Save current R image (workspace) as .RData file
print("Save R image (workspace) as .RData file named:")
f.image <- file.path(pathF.RDS, paste0("MiSeq_", run.date, "_preprocessing_image.RData"))
f.image
save.image(f.image)
```

```{r calculate-final-processing-time, results = "hold"}
# Calculate total processing time
print("Total processing time (seconds):")
proc.time() - start_time
```

```{r open R image, eval=FALSE, include=FALSE}
load("../results/00_dada2_output/RDS/MiSeq_2019_05_25_preprocessing_image.RData")
load("../results/00_dada2_output/RDS/MiSeq_2019_05_25_preprocessing_image.RData")
```
